Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-22 doc/mc-paper.tex
fix double reference
Reworded comparision to det-DNNF and dec-DNNF

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

@@

POGs generalize dec-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item As with det-DNNF~\cite{darwiche:aaai:2002,darwiche:jair:2002}, they allow arbitrary arguments to a $\por$ operation, as long as
  these have disjoint models.  By contrast, each
  sum node $\nodeu$ in a dec-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a \emph{decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating dec-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

We use \emph{reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
\emph{antecedent} clauses, such that each $D_i$ is in $\theta$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

Our proof generator is written in C/C++ and uses
\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
proof steps back into hinted CPOG clause additions, the generator can
use either its own RUP proof generator, or it can invoke
\dtrim{}~\cite{RAT}.  The latter yields shorter proofs and scales well
to large proofs, but each invocation has a high startup cost.  We
therefore only use it when solving larger problems (currently ones
with over 1000 clauses).

>>>>

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{heule:cade:2013,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

@@

We can now define and compare the two related representations:
\begin{itemize}
\item A det-DNNF graph is a POG with negation applied only to the variables.
\item A dec-DNNF graph is det-DNNF graph with the further restriction that any
  sum node $\nodeu$ has have exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating dec-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

We use \emph{reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{heule:cade:2013,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
\emph{antecedent} clauses, such that each $D_i$ is in $\theta$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

Our proof generator is written in C/C++ and uses
\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
proof steps back into hinted CPOG clause additions, the generator can
use either its own RUP proof generator, or it can invoke
\dtrim{}~\cite{heule:cade:2013}.  The latter yields shorter proofs and scales well
to large proofs, but each invocation has a high startup cost.  We
therefore only use it when solving larger problems (currently ones
with over 1000 clauses).
